---
title: "DTSA 5301 - COVID-19 Report"
author: "Ashish Bhutiani"
date: "6/20/2021"
output: pdf_document
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE,tidy.opts=list(width.cutoff=60), size="small")
library(tidyverse)
library(ggplot2)
library(lubridate)
```

# Question

Was there a bigger spike in cases in Miami/Dade county after the curfew was lifted versus other counties in Florida that didn't have the curfew?

# Data Source and Summary

In order to attempt to answer this question, we will be using a dataset provided by Johns Hopkins University and is an aggregation of data from multiple sources.

The URL for the repository holding the dataset is: https://github.com/CSSEGISandData/COVID-19

```{r dataload}
base_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv","time_series_covid19_deaths_US.csv")
urls <- str_c(base_url,file_names)
us_cases <- read.csv(urls[1])
us_deaths <- read.csv(urls[2])
```


# Data Cleanup

The information isn't exactly in the format we want to work with it in. We will perform the following data cleanup operations:

* Pivot Cases and Deaths to create a vertical vs horizontal dataset
* Join cases and deaths
* Convert the date column to a date type
* Filter on only records where cases are more than 0 and the state is Florida.

```{r data_clean}
pivoted_cases <- us_cases %>%
  pivot_longer(cols=starts_with('X'),
               names_to="date",
               names_prefix = "X",
               values_to="cases") %>%
  select(-c('Lat', 'Long_'))
pivoted_deaths <- us_deaths %>%
  pivot_longer(cols=starts_with('X'),
               names_to="date",
               names_prefix = "X",
               values_to="deaths") %>%
  select(-c('Lat', 'Long_','UID','iso2','iso3','code3','FIPS')) 
florida <- pivoted_cases %>%
  full_join(pivoted_deaths) %>%
  mutate(date = mdy(date)) %>%
  filter(cases > 0,Province_State=='Florida',Admin2 != "Unassigned")
summary(florida)
```

# Analysis and Visualization

Now let's look at what values we have in our data:

```{r head}
head(florida)
```

Let's summarize the data and see total cases/deaths by county:

```{r pie}
county_counts <- florida %>%
  group_by(Admin2) %>%
  summarize(cases=max(cases),deaths=max(deaths),population=max(Population)) %>%
  mutate(deaths_per_thousand = deaths * 1000 / population) %>%
  mutate(cases_per_thousand = cases * 1000 / population) %>%
  select(Admin2, cases, deaths, cases_per_thousand, deaths_per_thousand, population) 
ordered_county_counts <- county_counts[order(county_counts$cases),]
dotchart(ordered_county_counts$cases,main = "Cases by Florida Country",labels = ordered_county_counts$Admin2, cex=0.4)
```

We can see that Miami-Dade had the highest number of cases but if we look at the same data adjusted for population, then we will see a difference in the data.

```{r plot2}
ordered_county_counts <- county_counts[order(county_counts$cases_per_thousand),]
dotchart(ordered_county_counts$cases_per_thousand,main = "Cases (per 1000) by Florida Country",labels=ordered_county_counts$Admin2, cex=0.4)
```

This shows that we would want to use cases per 1000 as a more representative marker to compare to the rest of the state.

Let's also build a linear model.

```{r model}
linearMod <- lm(deaths ~ cases,data=florida)

summary(linearMod)
```

The Miami Dade curfew ended on April 12,2021 so we are interested in data after that point.

Let's just graph the total cases in Miami Dade after that date.

```{r miami_timechart}
miami_data <- florida %>%
  filter(Admin2=="Miami-Dade",date >= '2021-04-12')
ggplot(miami_data, aes(x=date, y=cases)) +
  geom_line() + 
  xlab("")
```

Now let's add the rest of Florida to the graph:
```{r timechart}
chart_data <- florida %>%
  filter(date >= '2021-04-12') %>%
  mutate(County=ifelse(Admin2=="Miami-Dade","Miami-Dade","Other")) %>%
  group_by(County,date) %>%
  summarize(cases=max(cases),deaths=max(deaths),population=max(Population)) %>%
  mutate(deaths_per_thousand = deaths * 1000 / population) %>%
  mutate(cases_per_thousand = cases * 1000 / population) %>%
  select(County,date, cases, deaths, cases_per_thousand, deaths_per_thousand, population) 
ggplot(chart_data, aes(x = date, y = cases_per_thousand)) + 
  geom_line(aes(color = County, linetype = County)) + 
  scale_color_manual(values = c("darkred", "steelblue")) +
  scale_y_log10()
```


# Conclusion and Bias
When we look at the data graphed together, we dont see a huge difference in the curves, so we can answer the question with No.

In terms of Bias, I thought that we would see a jump going in, but I didn't let it impact where I was going with the data. I think further analyis on the rate of change in cases might be warranted to see if there are any other anomalies.

# Session Info
```{r sessioninfo}
sessionInfo()
```